---
title: "Cervical Cancer Risk Prediction"
author: "Team 5 - ADS 503"
date: "2025-05-31"
output:
  pdf_document: default
  html_document: default
---

## Introduction

This project is a predictive modeling analysis focused on cervical cancer. The dataset was collected at Hospital Universitario de Caracas in Venezuela and includes patient demographic, lifestyle, and medical history information. The goal is to build models that can predict whether a patient is likely to test positive for cervical cancer based on biopsy outcomes.

```{r setup, include=FALSE}
# Load libraries
library(tidyverse)
library(caret)
library(janitor)
library(skimr)
library(purrr)
library(GGally)
library(corrplot)
```

## Data Importing and Pre-Processing
```{r}
cervical_data <- read_csv("risk_factors_cervical_cancer.csv")
```

```{r}
head(cervical_data, 10)
View(head(cervical_data, 10))
```

```{r}
nrow(cervical_data)  # Number of rows (patients)
ncol(cervical_data)  # Number of columns (features)
```

```{r}
# cleaning up feature names
cervical_data <- cervical_data %>% clean_names()

# need to manually rename a few features
cervical_data <- cervical_data %>%
  rename(
    stds = st_ds,
    stds_number = st_ds_number,
    stds_condylomatosis = st_ds_condylomatosis,
    stds_cervical_condylomatosis = st_ds_cervical_condylomatosis,
    stds_vaginal_condylomatosis = st_ds_vaginal_condylomatosis,
    stds_vulvo_perineal_condylomatosis = st_ds_vulvo_perineal_condylomatosis,
    stds_syphilis = st_ds_syphilis,
    stds_pelvic_inflammatory_disease = st_ds_pelvic_inflammatory_disease,
    stds_genital_herpes = st_ds_genital_herpes,
    stds_molluscum_contagiosum = st_ds_molluscum_contagiosum,
    stds_aids = st_ds_aids,
    stds_hiv = st_ds_hiv,
    stds_hepatitis_b = st_ds_hepatitis_b,
    stds_hpv = st_ds_hpv,
    stds_number_of_diagnosis = st_ds_number_of_diagnosis,
    stds_time_since_first_diagnosis = st_ds_time_since_first_diagnosis,
    stds_time_since_last_diagnosis = st_ds_time_since_last_diagnosis
  )

colnames(cervical_data)
```

```{r}
# Convert ? to NA
cervical_data[cervical_data == "?"] <- NA
sum(cervical_data == "?", na.rm = TRUE)

```

```{r}
# Show unique values for each column
map(cervical_data, ~ unique(.) %>% sort()) %>%
  enframe(name = "column", value = "unique_values") %>%
  print(n = Inf)
```

```{r}
# convert all character columns to numeric
cervical_data <- cervical_data %>%
  mutate(across(where(is.character), ~ as.numeric(.)))

# recategorizing binary indicator variables as categorical (factor) type
# note: All binary variables are coded as 0 = “No” and 1 = “Yes”.
factor_vars <- c("smokes", "hormonal_contraceptives", "iud", "stds",
                 "stds_condylomatosis", "stds_cervical_condylomatosis", 
                 "stds_vaginal_condylomatosis", "stds_vulvo_perineal_condylomatosis",
                 "stds_syphilis", "stds_pelvic_inflammatory_disease", 
                 "stds_genital_herpes", "stds_molluscum_contagiosum", 
                 "stds_aids", "stds_hiv", "stds_hepatitis_b", "stds_hpv",
                 "dx_cancer", "dx_cin", "dx_hpv", "dx", 
                 "hinselmann", "schiller", "citology", "biopsy")

cervical_data <- cervical_data %>%
  mutate(across(all_of(factor_vars), ~ as.factor(.)))
```

```{r}
head(cervical_data, 10)
View(head(cervical_data, 10))
```

```{r}
# view missing data
colSums(is.na(cervical_data))
```

```{r}
# drop sparse columns with >90% missing values (too incomplete for modeling)
cervical_data <- cervical_data %>%
  select(-stds_time_since_first_diagnosis, -stds_time_since_last_diagnosis)
```

```{r}
# Select numeric columns and remove rows with NAs temporarily
numeric_data <- cervical_data %>%
  select(where(is.numeric)) %>%
  drop_na()

# Compute correlation matrix
cor_matrix <- cor(numeric_data)

# Plot correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.8)
```

```{r}
# Missing Value Imputation

# separate numeric columns
numeric_vars <- cervical_data %>%
  select(where(is.numeric))

# apply median imputation
preproc <- preProcess(numeric_vars, method = "medianImpute")
numeric_imputed <- predict(preproc, numeric_vars)

# recombine with non-numeric columns (factors)
non_numeric <- cervical_data %>%
  select(where(Negate(is.numeric)))

# Final imputed dataset, and saving separately for models that require no NAs for modeling
cervical_data_imputed <- bind_cols(numeric_imputed, non_numeric)
```

```{r}
# Create binary numeric version of biopsy (0 = No, 1 = Yes)
cervical_data_imputed$biopsy_numeric <- as.numeric(cervical_data_imputed$biopsy) - 1

# Select numeric predictors (exclude biopsy_numeric itself)
numeric_vars <- cervical_data_imputed %>%
  select(where(is.numeric)) %>%
  select(-biopsy_numeric)

# Compute point-biserial correlation between each predictor and biopsy_numeric
cor_results <- sapply(numeric_vars, function(x) {
  cor.test(cervical_data_imputed$biopsy_numeric, x)$estimate
})

# Display sorted correlations from strongest to weakest
sort(cor_results, decreasing = TRUE)
```
The correlation is all very weak, all less than 0.1.




## Exploratory Data Analysis

```{r}
summary(cervical_data)
```

```{r}
# drop features with only one unique value and NA: no predictive power
cervical_data <- cervical_data %>%
  select(-stds_cervical_condylomatosis, -stds_aids)
```
We found that stds_cervical_condylomatosis and stds_aids each contained only one unique non-missing value (all "0", or "No") and had a high proportion of missing values (NA). This indicates that they provide no meaningful variation for modeling and would add unnecessary sparsity to the data. While it’s possible that the missingness is related to the value itself (e.g., respondents choosing not to disclose due to sensitivity), the lack of variation in the observed data makes it impossible to model these features reliably. This aligns with guidance that missingness dependent on the unobserved value itself (i.e., Not Missing At Random) presents a particularly difficult modeling scenario (Kuhn & Johnson, 2013). **Citing WEEK 2 Discussion post reading, include in final paper!

```{r}
# we want to do more discovery in selecting the target variable
table(cervical_data$biopsy)
table(cervical_data$dx_cancer)
```
This table tells us that biopsy positives are more common than positive cancer diagnoses. Only 18 patients were diagnosed with cancer, which is very small for a classification target.

```{r}
tab <- table(cervical_data$biopsy, cervical_data$dx_cancer)

dimnames(tab) <- list(
  Biopsy = c("Negative", "Positive"),
  CancerDiagnosis = c("No Cancer", "Cancer")
)
addmargins(tab)
```
To note from this table: Not all positive biopsies were diagnosed as cancer. 49 patients had a positive biopsy, but no confirmed cancer diagnosis (could be precancerous). Some patients were diagnosed with cancer despite a negative biopsy (could be a preexisting diagnosis).

```{r}
ggplot(cervical_data, aes(x = biopsy, fill = dx_cancer)) +
  geom_bar(position = "dodge") +
  labs(title = "Biopsy Results by Cancer Diagnosis",
       x = "Biopsy Result", y = "Count", fill = "Cancer Diagnosis")
```

```{r}
ggplot(cervical_data, aes(x = age)) + 
  geom_histogram(binwidth = 5, fill = "blue") +
  labs(title = "Age Distribution")
```
The age distribution of patients is right-skewed, with the majority of patients between 20 and 40 years old. A small number of patients are over 50, and very few are over 70. This suggests that the dataset largely reflects a younger population, which aligns with the typical age range for cervical cancer screening. However, outliers in the older range could be important to monitor for elevated risk patterns.


```{r}
ggplot(cervical_data, aes(x = biopsy, y = age)) +
  geom_boxplot() +
  labs(title = "Age by Biopsy Outcome")
```
The median age for patients with a positive biopsy result appears slightly higher than for those with a negative result. While there is considerable overlap in the distributions, the boxplot suggests that older patients may be more likely to test positive for cervical cancer. A few older individuals with negative biopsy results appear as outliers, but the positive group shows a more concentrated distribution between ages 25 and 45.

```{r}
ggplot(cervical_data, aes(x = smokes, fill = biopsy)) +
  geom_bar(position = "fill") +
  labs(title = "Smoking Status by Biopsy Result")
```
This bar plot shows the proportion of biopsy outcomes (0 = Negative, 1 = Positive) across smoking status groups (0 = non-smoker, 1= smoker, NA = missing value). The distribution of biopsy results appears very similar across all three smoking categories. This suggests that smoking status does not show a strong relationship with biopsy outcome in this dataset.

```{r}
# view the distribution of years of hormonal contraceptive use across biopsy (using the imputed dataset to avoid NAs)
ggplot(cervical_data_imputed, aes(x = biopsy, y = hormonal_contraceptives_years)) +
  geom_boxplot() +
  labs(title = "Hormonal Contraceptive Use (Years) by Biopsy Result")
```
Although the medians for hormonal contraceptive use are similar between biopsy outcome groups, the distribution for the positive biopsy group is more spread out and contains more patients with slightly longer years of use. This could indicate that patients with positive biopsy results may have a few more years of contraceptive use (although correlation was weak).


```{r}


```


